Multi-Label Speech Emotion Recognition Using 2D Convolutional Neural Networks

### 2019-08-30 research/paper/paper.tex
Complete paper draft

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{dblfloatfix}
\usepackage{caption}
\graphicspath{{figures/}}
\usepackage[strings]{underscore}
\usepackage{textcomp}
\usepackage{multirow}
\usepackage[table,xcdraw]{xcolor}
\usepackage{pifont}
\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08emT\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

@@

\author{\IEEEauthorblockN{Brian Pho, Thomas Truong, Svetlana Yanushkevich}
\IEEEauthorblockA{\textit{Biometric Technologies Laboratory, Department of Electrical and Computer Engineering}\\
University of Calgary, Canada\\\{brian.pho, thomas.truong, syanshk\}@ucalgary.ca}}

@@

\begin{abstract}
Current speech emotion recognition systems often overlook the multi-label data that comes with databases. In this paper, we address the problem of whether machine learning can be used to detect multiple emotions in speech. We created a combined database consisting of four voice emotion-labeled databases, and we used it to train a 2D convolutional neural network to determine if the model could recognize multiple emotions in a speech sample. The model was able to classify the samples with an accuracy of 52.57\%. This result shows that it is possible to apply machine learning to the problem of multi-label speech emotion recognition and to achieve a reasonable accuracy.
\end{abstract}

@@

The field of affective computing studies the development of systems that can recognize, interpret, process, and simulate human affects. A subfield of affective computing that we are interested in is speech emotion recognition (SER). SER is the problem of recognizing which emotions are present in speech and SER has been approached in two ways:
\begin{itemize}
	\item Feature engineering
	\item Machine learning (ML)
\end{itemize} 
Feature engineering is a method of manually extracting desired parameters from the data for use in classification and recognition. For audio, examples of parameters include pitch, energy, and Mel-Frequency Cepstral Coefficients (MFCC).\cite{Rybka2013} Nassif et al. \cite{Nassif2019} found that most researchers used the MFCC parameter for deep learning models but also recommended other parameters such as Linear Predictive Coding.

Machine learning is a method of creating an artificial neural network that automatically extracts its own parameters for classification and recognition. Examples include support vector machines (SVM), convolutional neural networks (CNN), and recurrent neural networks (RNN). Both approaches are not exclusive and a hybrid approach of feature engineering and machine learning results in better classification accuracy.\cite{Nassif2019}

A common example of combining both approaches is to convert an audio waveform into a spectrogram using the short-time Fourier transform (STFT) and to feed the spectrogram into a neural network. Papers such as \cite{Engel2019}, \cite{Chen2018}, \cite{Badshah2019}, and \cite{Zhao2019} have demonstrated that this approach is successful for processing audio and for recognizing emotions in speech.

Current state-of-the-art speech emotion recognition has achieved an accuracy of 52.14\% for the case of speaker-independent, single-labeled emotions on the IEMOCAP database.\cite{Zhao2019} This result was achieved using a 2D CNN LSTM model where the audio waveform was converted into a log-Mel spectrogram and then fed into the model.

@@

We approach the problem of multi-label emotion classification by using a 2D CNN to classify speech samples into multiple emotions. We collected four databases: two with multi-label and single-label samples, and two with only single-label samples. We then combined the four databases into an a single database by processing all of the speech samples into log-Mel spectrograms. These log-Mel spectrograms were then fed into a seven layer neural network consisting of four convolutional layers and three dense layers. After training the neural network, the model achieves a 52.57\% test accuracy.

@@

\bgroup
\def\arraystretch{2}%  1 is the default, change whatever you need
\begin{table*}
	\resizebox{\textwidth}{!}{%
		\begin{tabular}{|l|c|c|c|c|c|c|c|c|c|c|c|c|c|}
			\hline
			\multicolumn{1}{|c|}{} & \multicolumn{13}{c|}{\textbf{Emotions}} \\ \cline{2-14} 
			\multicolumn{1}{|c|}{\multirow{-2}{*}{\textbf{Databases}}} & \textbf{Neutral} & \textbf{Anger} & \textbf{Disgust} & \textbf{Fear} & \textbf{Happy} & \textbf{Sad} & \textbf{Surprise} & \textbf{Calm} & \textbf{Excitement} & \textbf{Frustration} & \textbf{Amused} & \textbf{Sleepy} & \textbf{Bored} \\ \hline
			\rowcolor[HTML]{F6F8FA} 
			\textbf{IEMOCAP} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark \\ \hline
			\textbf{CREMA-D} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ \hline
			\rowcolor[HTML]{F6F8FA} 
			\textbf{TESS} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ \hline
			\textbf{RAVDESS} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ \hline
			\rowcolor[HTML]{F6F8FA} 
			\textbf{MSP-IMPROV} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ \hline
			\textbf{SAVEE} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ \hline
			\rowcolor[HTML]{F6F8FA} 
			\textbf{Emo-DB} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark \\ \hline
			\textbf{EmoV-DB} & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark & \cmark & \xmark \\ \hline
		\end{tabular}%
	}
	\caption{Comparison of databases with their labeled emotions.}
	\label{tab:db_emo_table}
\end{table*}
\egroup

@@

To determine if a ML model can perform multi-label speech emotion recognition, we first need to obtain speech samples with their labeled emotion. We considered eight databases and chose four based on accessibility and the number of overlapping emotions. Table \ref{tab:db_emo_table} compares the set of emotions of each database which makes it easier to identify overlapping emotions. We chose four databases because it becomes more difficult to maintain consistency due to database variability. Databases can vary in
\begin{itemize}
	\item The set and number of labeled emotions
	\item The number of labels per sample
	\item The audio quality such as sampling rate and noise
	\item The spoken language
\end{itemize}
Given this variability, we chose the following four databases
\begin{enumerate}
	\item IEMOCAP \cite{busso_2008}
	\item TESS \cite{dupuis_2011}
	\item RAVDESS \cite{livingstone_2018}
	\item CREMA-D \cite{cao_2014}
\end{enumerate}

To control for the set and number of labeled emotions, we consider the following seven emotions for recognition
\begin{enumerate}
	\item Neutral
	\item Anger
	\item Disgust
	\item Fear
	\item Happy
	\item Sad
	\item Surprise
\end{enumerate}
We chose these seven emotions due to them being considered basic emotions by \cite{Ekman1992} and due to these seven being the most common among all databases.

@@

We combined these four databases into a combined database to train, validate, and test the neural network. The combined database is detailed in figure \ref{fig:combined_db}.

\begin{figure*}[t]
	\centering
	\hspace{6mm}
	\includegraphics[width=\textwidth]{combined_db.png} 
	\caption{Proportions of each: database, emotion, and label types in the combined database.}
	\label{fig:combined_db}
\end{figure*}

Each sample from a database would flow through the same preprocessing steps to maintain consistency across all samples. A high level overview of the preprocessing steps is shown in figure \ref{fig:high_level_dataflow_diagram} and the detailed steps are described below.
\begin{enumerate}
	\item A sample starts as a raw waveform in the form of time series points specifying the amplitude at a certain point in time.
	\item The sample is then padded or cropped to the desired length of 4.5s. Shorter samples were zero-padded on the right tail. Longer samples were cropped and the extra information discarded.
	\item If the sample came from a database that we considered noisy, then a noise reduction filter was applied to the sample. We consider the IEMOCAP and CREMA-D databases to be noisy.
	\item The sample is then converted into a log-Mel spectrogram using the short-time Fourier transform (STFT) and Mel scale equation. The phase information was discard as it does not seem to hold relevant information. \cite{Kozakowski2017}
	\item The final step is to normalize the spectrograms to have values between negative one and one. This was done by using a minmax scaling function.
\end{enumerate}

@@

\begin{figure}[]
	\centering
	\hspace{6mm}
	\includegraphics[width=\linewidth]{high_level_dataflow_diagram.png}
	\caption{A high level overview of the processing stages that a speech sample goes through.}
	\label{fig:high_level_dataflow_diagram}
\end{figure}

@@

\begin{figure*}[b!]
	\centering
	\hspace{6mm}
	\includegraphics[width=\textwidth]{neural_network_architecture.png}
	\caption{Architecture for the speech emotion recognition network including the input layer and seven processing layers.}
	\label{fig:neural_network_architecture}
\end{figure*}

After all of the databases were processed this way, the final combined database was fed into a neural network for training, validation, and testing. We constructed a seven layer neural network consisting of four convolutional layers and three dense layers. The network architecture is shown in figure \ref{fig:neural_network_architecture}. The training process is described below.
\begin{itemize}
	\item We shuffled the combined database to make each input batch more uniform thus mitigating large gradient updates.
	\item We split the combined database into 80\% training, 10\% validation, and 10\% testing.
	\item We applied dropout \cite{Srivastava2014} to the dense layers and batch normalization \cite{Ioffe2015} to the convolutional layers to deal with overfitting.
	\item We updated the model's hyperparameters based on the validation loss and accuracy to improve the model's accuracy and ability to generalize.
	\item We used class weights during training to address class imbalance.
\end{itemize}

Table \ref{tab:model_hyperparams} describes the network hyperparameters and table \ref{tab:train_hyperparams} describes the training hyperparameters.

\begin{table}[b]
	\resizebox{\linewidth}{!}{%
		\begin{tabular}{|l|r|}
			\hline
			\textbf{Hyperparameter} & \textbf{Value} \\ \hline
			\rowcolor[HTML]{F6F8FA} 
			Input Dimensions & 278w x 200h x 1d \\ \hline
			Optimization Algorithm & Adam \\ \hline
			\rowcolor[HTML]{F6F8FA} 
			Loss Measure & Binary Cross-entropy \\ \hline
			Accuracy Metric & Categorical Cross-entropy \\ \hline
			\rowcolor[HTML]{F6F8FA} 
			Activation Function & Sigmoid \\ \hline
		\end{tabular}%
	}
	\caption{The model's hyperparameters.}
	\label{tab:model_hyperparams}
\end{table}

\begin{table}[]
	\resizebox{\linewidth}{!}{%
		\begin{tabular}{|l|r|}
			\hline
			\textbf{Hyperparameter} & \textbf{Value} \\ \hline
			\rowcolor[HTML]{F6F8FA} 
			Epochs & 20 \\ \hline
			Batch Size & 32 \\ \hline
			\rowcolor[HTML]{F6F8FA} 
			Training Set & 17,341 \\ \hline
			Validation Set & 2,167 \\ \hline
			\rowcolor[HTML]{F6F8FA} 
			Testing Set & 2,167 \\ \hline
		\end{tabular}%
	}
	\caption{The training hyperparameters.}
	\label{tab:train_hyperparams}
\end{table}

@@

After training the model, the model was evaluated on the testing set to get the final accuracy of the model. The final accuracy achieved was 52.57\%. Confusion matrices for each emotion are shown in Figure \ref{fig:confusion_matrix}.

\begin{figure}[]
	\centering
	\hspace{6mm}
	\includegraphics[width=\textwidth]{confusion_matrix.png}
	\caption{Confusion matrices for each emotion.}
	\label{fig:confusion_matrix}
\end{figure}

@@

We created a neural network model that achieved a classification accuracy of 52.57\% on the problem of multi-label emotion classification. This model can successfully classify multiple emotions in a sample and while this is a good result, we can do better.

@@

One limitation of this work was that we only accounted for seven emotions but we know that there are more emotions than this. \cite{} Another limitations was that 
Here we present future improvements of our work 
\begin{itemize}
	\item Improving the recognition accuracy by using more sophisticated neural network architectures such as LSTMs or using more databases.
	\item Incorporating the phase data from the STFT and testing to see if that improves the accuracy.
	\item Testing a binary relevant approach to this multi-label problem and comparing it to this model.
\end{itemize}

@@

Overall, this paper presented a convolutional neural network model that achieves a 52.57\% accuracy on multi-class, multi-label speech emotion recognition using four combined databases. We obtained this result by transforming raw speech samples into log-Mel spectrograms using STFT and the Mel scale. These log-Mel spectrograms are then fed into a eight layer neural network for classification for training. When we have completed training the neural network, the model was evaluated and achieved as 52.57\% accuracy. While this result is a promising start, we suggest improvements that future work can build upon to improve the accuracy.

\section*{Acknowledgment}
Acknowledge PURE award.

>>>>

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{dblfloatfix}
\usepackage{hyperref}
\usepackage{caption}
\usepackage{pifont}
\usepackage{textcomp}
\usepackage{multirow}
\usepackage[table,xcdraw]{xcolor}
\usepackage[strings]{underscore}

\graphicspath{{figures/}}
\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08emT\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

@@

\author{\IEEEauthorblockN{Brian Pho, Thomas Truong, Svetlana Yanushkevich}
\IEEEauthorblockA{\textit{Biometric Technologies Laboratory, Department of Electrical and Computer Engineering}\\
University of Calgary, Canada\\
\{brian.pho, thomas.truong, syanshk\}@ucalgary.ca}}

@@

\begin{abstract}
Current speech emotion recognition systems often overlook the multi-label data that comes with databases. In this paper, we address the problem of whether machine learning can be used to detect multiple emotions in speech. We created a combined database consisting of four speech emotion-labeled databases, and we used it to train a 2D convolutional neural network to determine if the model could recognize multiple emotions in a speech sample. The model was able to classify the samples with an accuracy of 52.57\%. This shows that it is possible to apply machine learning to the problem of multi-label speech emotion recognition and to achieve a reasonable accuracy.\footnote{\url{https://github.com/Brian-Pho/RVST598_Speech-Emotion-Recognition}}
\end{abstract}

@@

The field of affective computing studies the development of systems that can recognize, interpret, process, and simulate human affects. A subfield of affective computing that we are interested in is speech emotion recognition (SER). SER is the problem of recognizing which emotions are present in speech and SER has been approached in two ways:
\begin{itemize}
	\item Feature engineering
	\item Machine learning (ML)
\end{itemize} 
Feature engineering is a method of manually extracting desired parameters from the data for use in  recognition. For audio, examples of parameters include pitch, energy, and Mel-Frequency Cepstral Coefficients (MFCC).\cite{Rybka2013} Nassif et al. \cite{Nassif2019} found that most researchers used the MFCC parameter for deep learning models but also recommended other parameters such as Linear Predictive Coding.

Machine learning is a method of creating an artificial neural network that automatically extracts its own parameters for recognition. Examples include support vector machines (SVM), convolutional neural networks (CNN), and recurrent neural networks (RNN). Both approaches are not exclusive and a hybrid approach of feature engineering and machine learning results in better classification accuracy.\cite{Nassif2019}

A common example of combining both approaches is to convert an audio waveform into a spectrogram using the Short-Time Fourier Transform (STFT) and to feed the spectrogram into a neural network. Papers such as \cite{Engel2019}, \cite{Chen2018}, \cite{Badshah2019}, and \cite{Zhao2019} have demonstrated that this approach is successful for processing audio and for recognizing emotions in speech.

Current state-of-the-art speech emotion recognition has achieved an accuracy of 52.14\% for the case of speaker-independent, single-labeled emotions on the IEMOCAP database.\cite{Zhao2019} This result was achieved using a 2D CNN long short-term memory (LSTM) model where the audio waveform was converted into a log-Mel spectrogram and then fed into the model.

@@

We approach the problem of multi-label emotion classification by using a 2D CNN to classify speech samples into multiple emotions. We collected four databases: two with multi-label and single-label samples, and two with only single-label samples. We then combined the four databases into an a single database by processing all of the speech samples into log-Mel spectrograms. These log-Mel spectrograms were then fed into an eight-layer neural network consisting of four convolutional layers and four dense layers. After training the neural network, the model achieves a 52.57\% test accuracy.

@@

\bgroup
\def\arraystretch{2}
\begin{table*}
	\resizebox{\textwidth}{!}{%
		\begin{tabular}{|l|c|c|c|c|c|c|c|c|c|c|c|c|c|}
			\hline
			\multicolumn{1}{|c|}{} & \multicolumn{13}{c|}{\textbf{Emotions}} \\ \cline{2-14} 
			\multicolumn{1}{|c|}{\multirow{-2}{*}{\textbf{Databases}}} & \textbf{Neutral} & \textbf{Anger} & \textbf{Disgust} & \textbf{Fear} & \textbf{Happy} & \textbf{Sad} & \textbf{Surprise} & \textbf{Calm} & \textbf{Excitement} & \textbf{Frustration} & \textbf{Amused} & \textbf{Sleepy} & \textbf{Bored} \\ \hline
			\rowcolor[HTML]{F6F8FA} 
			\textbf{IEMOCAP} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark \\ \hline
			\textbf{CREMA-D} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ \hline
			\rowcolor[HTML]{F6F8FA} 
			\textbf{TESS} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ \hline
			\textbf{RAVDESS} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ \hline
			\rowcolor[HTML]{F6F8FA} 
			\textbf{MSP-IMPROV} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ \hline
			\textbf{SAVEE} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ \hline
			\rowcolor[HTML]{F6F8FA} 
			\textbf{Emo-DB} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark \\ \hline
			\textbf{EmoV-DB} & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark & \cmark & \xmark \\ \hline
		\end{tabular}%
	}
	\caption{Comparison of databases with their labeled emotions.}
	\label{dbEmoTable}
\end{table*}
\egroup

@@

To determine if a ML model can perform multi-label speech emotion recognition, we first need to obtain speech samples with their labeled emotion. We considered eight databases and chose four based on accessibility and the number of overlapping emotions. Table \ref{dbEmoTable} compares the set of emotions of each database which makes it easier to identify overlapping emotions. We chose four databases because it becomes more difficult to maintain consistency due to database variability. Databases can vary in
\begin{itemize}
	\item The set and number of labeled emotions
	\item The number of labels per sample
	\item The audio quality such as sampling rate and noise
	\item The spoken language
\end{itemize}
Given this variability, we chose the following four databases
\begin{enumerate}
	\item IEMOCAP \cite{busso_2008}
	\item TESS \cite{dupuis_2011}
	\item RAVDESS \cite{livingstone_2018}
	\item CREMA-D \cite{cao_2014}
\end{enumerate}

To control for the set and number of labeled emotions, we consider the following seven emotions for recognition
\begin{enumerate}
	\item Neutral
	\item Anger
	\item Disgust
	\item Fear
	\item Happy
	\item Sad
	\item Surprise
\end{enumerate}
We chose these seven emotions due to them being considered basic emotions by Ekman \cite{Ekman1992} and due to these seven being the most common among all databases.

@@

We combined these four databases into a combined database to train, validate, and test the neural network. The combined database is detailed in figure \ref{combinedDb}.

\begin{figure*}[t]
	\centering
	\hspace{6mm}
	\includegraphics[width=\textwidth]{combined_db.png} 
	\caption{Proportions of each database, emotion, and label types in the combined database.}
	\label{combinedDb}
\end{figure*}

Each sample from a database would flow through the same preprocessing steps to maintain consistency across all samples. A high level overview of the preprocessing steps is shown in figure \ref{highLevelDataFlowDiagram} and the detailed steps are described below.
\begin{enumerate}
	\item A sample starts as a raw waveform in the form of time series points specifying the amplitude at a certain point in time.
	\item The sample is then padded or cropped to the desired length of 4.5s. Shorter samples were zero-padded on the right tail. Longer samples were cropped and the extra information discarded.
	\item If the sample came from a database that we considered noisy, then a noise reduction filter was applied to the sample. We consider the IEMOCAP and CREMA-D databases to be noisy.
	\item The sample is then converted into a log-Mel spectrogram using the short-time Fourier transform (STFT) and Mel scale equation. The phase information was discard as it does not seem to hold relevant information. \cite{Kozakowski2017}
	\item The final step is to normalize the spectrograms to have values between negative one and one. This was done by using a min-max scaling function.
\end{enumerate}

The discrete-time STFT is defined in (\ref{stft}) where $x[n]$ is the discrete signal and $w[n]$ is the window function.

\begin{equation}
	\label{stft}
	\mathbf{STFT}\{x[n]\}(m,\omega) \equiv X(m,\omega) = \sum_{n=-\infty}^{\infty}x[n]w[n-m]e^{-j \omega n}
\end{equation}

The Mel scale is defined in (\ref{melScale}) where $f$ is the frequency in hertz and $m$ is mels.\cite{Oshaughnessy1990}

\begin{equation}
\label{melScale}
m = 2595 log_{10}(1 + \frac{f}{700})
\end{equation}

@@

\begin{figure}[]
	\centering
	\hspace{6mm}
	\includegraphics[width=\linewidth]{high_level_dataflow_diagram.png}
	\caption{A high level overview of the processing stages that a speech sample goes through.}
	\label{highLevelDataFlowDiagram}
\end{figure}

@@

\begin{figure*}[h!]
	\centering
	\hspace{6mm}
	\includegraphics[width=\textwidth]{neural_network_architecture.png}
	\caption{Architecture for the speech emotion recognition network including the input layer and eight processing layers.}
	\label{neuralNetworkArchitecture}
\end{figure*}

After all of the databases were processed this way, the final combined database was fed into a neural network for training, validation, and testing. We constructed a eight layer neural network consisting of four convolutional layers and three dense layers. The network architecture is shown in figure \ref{neuralNetworkArchitecture}. The training process is described below.
\begin{itemize}
	\item We shuffled the combined database to make each input batch more uniform thus mitigating large gradient updates.
	\item We split the combined database into 80\% training, 10\% validation, and 10\% testing.
	\item We applied dropout to the dense layers and batch normalization to the convolutional layers to deal with overfitting. \cite{Srivastava2014}, \cite{Ioffe2015}
	\item We updated the model's hyperparameters based on the validation loss and accuracy to improve the model's accuracy and ability to generalize.
	\item We used class weights during training to address class imbalance.
\end{itemize}

Table \ref{modelHyperparams} describes the network hyperparameters and table \ref{trainHyperparams} describes the training hyperparameters.

\begin{table}[h!]
	\caption{The model's hyperparameters.}
	\label{modelHyperparams}
	\resizebox{\linewidth}{!}{%
		\begin{tabular}{|l|r|}
			\hline
			\textbf{Hyperparameter} & \textbf{Value} \\ \hline
			\rowcolor[HTML]{F6F8FA} 
			Input Dimensions & 278w x 200h x 1d \\ \hline
			Optimization Algorithm & Adam \\ \hline
			\rowcolor[HTML]{F6F8FA} 
			Loss Measure & Binary Cross-entropy \\ \hline
			Accuracy Metric & Categorical Cross-entropy \\ \hline
			\rowcolor[HTML]{F6F8FA} 
			Activation Function & Sigmoid \\ \hline
		\end{tabular}%
	}
\end{table}

\begin{table}[h!]
	\caption{The training hyperparameters.}
	\label{trainHyperparams}
	\resizebox{\linewidth}{!}{%
		\begin{tabular}{|l|r|}
			\hline
			\textbf{Hyperparameter} & \textbf{Value} \\ \hline
			\rowcolor[HTML]{F6F8FA} 
			Epochs & 20 \\ \hline
			Batch Size & 32 \\ \hline
			\rowcolor[HTML]{F6F8FA} 
			Training Set & 17,341 \\ \hline
			Validation Set & 2,167 \\ \hline
			\rowcolor[HTML]{F6F8FA} 
			Testing Set & 2,167 \\ \hline
		\end{tabular}%
	}
\end{table}

@@

After training the model, it was evaluated on the testing set to get the final accuracy. The final accuracy achieved was 52.57\%. Confusion matrices for each emotion are shown in Figure \ref{confusionMatrix}.

\begin{figure*}[h!]
	\centering
	\hspace{6mm}
	\includegraphics[width=\textwidth]{confusion_matrix.png}
	\caption{Confusion matrices for each emotion.}
	\label{confusionMatrix}
\end{figure*}

@@

We created a neural network model that achieved a 52.57\% accuracy on the problem of multi-label speech emotion classification. This is better than the state-of-the-art result of 52.14\% achieved in the single-label case. \cite{Zhao2019} So we conclude that it is possible to achieve a reasonable accuracy on the problem of multi-label speech emotion recognition by applying deep learning to the problem.

In analyzing the confusion matrices, we see that the model predicts an emotion is absent in most of the samples with "surprise" being the exception. We suspect that "surprise" being an exception is due to the class imbalance that is shown in figure \ref{combinedDb}. The "surprise" class was the least represented class out of the seven classes and this makes the model predict it more due to the use of class weights. Class weights bias the model towards underrepresented classes in an attempt to balance classes but this seems to have affected the model's ability to learn.

@@

One limitation of this work is that we only accounted for seven emotions but recent research has suggested that there are more emotions such as boredom, shame, and triumph. \cite{Cordaro2018} However, one issue with expanding the set of considered emotions is the lack of databases with the labeled emotion.

Another limitation of this work is that all samples are spoken in English so the model is biased towards Anglophones. In theory, the basic emotions are universal across languages and cultures so incorporating databases spoken in different languages, such as the Emo-DB database, would help the model generalize across languages.\cite{Burkhardt2005}

The following list is a suggestion that future work could pursue:
\begin{itemize}
	\item Using more sophisticated neural network architectures such as LSTMs or using more databases.
	\item Incorporating the phase data from the STFT.
	\item Testing a binary relevance approach to this multi-label problem.
	\item Replacing the use of STFT with a wavelet transform.
\end{itemize}

@@

Overall, this paper presented a 2D CNN model that achieved an accuracy of 52.57\% on the problem of multi-label speech emotion recognition using four combined databases. We obtained this result by transforming raw speech samples into log-Mel spectrograms using the STFT and the Mel scale. The log-Mel spectrograms are then fed into an eight layer neural network for classification. While this result is a promising start, we suggest improvements that future work can build upon to improve the accuracy of the model and to expand the scope of emotions considered.

\section*{Acknowledgment}
This work was funded by the Program for Undergraduate Research Experience (PURE) award granted by the University of Calgary.
